---
title: "The {datardis} package"
subtitle: "Explore the Dr Who TV series"
author: "Jonathan Kitt"
date: "2023-06-23"
toc: true
categories: [R, package, Dr Who]
---

# Install the package

<br>

The {datardis} package is available on [CRAN](https://cran.r-project.org/web/packages/datardis/index.html):

```{r, eval=FALSE}
install.packages("datardis")
```

<br>

# The datasets

<br>

As of today, the {datardis} package (version 0.0.3) contains six datasets :

-   List of **Doctor Who episodes** from the revived era (since 2005)

-   List of **directors** for Doctor Who episodes

-   List of **writers** for Doctor Who episodes

-   List of **Torchwood episodes**

-   List of **directors** for Torchwood episodes

-   List of **writers** for Torchwood episodes

<br>

## Accessing the datasets

<br>

To access the datasets, use the following commands:

```{r, eval=FALSE}
# Load the package
library(datardis)

# List of Dr Who episodes
drwho_episodes

# List of Dr Who directors
drwho_directors

# List of Dr Who writers
drwho_writers

# List of Torchwood episodes
torchwood_episodes

# List of Torchwood directors
torchwood_directors

# List of Torchwood writers
torchwood_writers

```

<br>

# Exploring the datasets

<br>

## List of Dr Who episodes

<br>

To view the dataset content, use the `glimpse()` function:

```{r, eval=FALSE}
# Load the tidyverse
library(tidyverse)

# View the dataset content
glimpse(drwho_episodes)
```

```{r, echo=FALSE, eval=TRUE}
dplyr::glimpse(datardis::drwho_episodes)
```

<br>

## Who wrote the most Dr Who episodes?

<br>

To find out who wrote the most Dr Who episodes, use the following code:

```{r, eval=FALSE}
drwho_writers |> 
  count(writer, sort = TRUE) |> 
  head(5)
```

```{r, echo=FALSE, eval=TRUE}
datardis::drwho_writers |> 
  dplyr::count(writer, sort = TRUE) |> 
  head(5)
```
